---
pagetitle: "calcofi4r"
output:
  github_document:
    html_preview: false
editor_options: 
  chunk_output_type: console
---

# calcofi4r <img src="man/figures/logo.svg" style="float:right; height:150px">

This R package contains functions for reading and visualizing data from the CalCOFI API at [api.calcofi.io](https://api.calcofi.io).

```{r knitr, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = F,
  comment = "#>",
  fig.path="man/figures/"
)
# devtools::document()
# devtools::install_local(here::here(), force=T) # install local
```

## Install

```{r, include = FALSE}
# install.packages("devtools")
# remotes::install_github("calcofi/calcofi4r")
devtools::load_all()
```

This package lives on Github, not yet CRAN, so you’ll need to run the
following to install or update the package:

```r
remotes::install_github("calcofi/calcofi4r")
```

Then to use the functions (and datasets) in the package:

```r
library(calcofi4r)
```

Please visit the vignette to [Get started](./articles/calcofi4r.html) and [Reference](reference/index.html) of functions for more.

## Code of Conduct

This is an open-source project so your input is greatly welcomed! Please note that the `calcofi4r` project is released with a [Contributor Code of Conduct](https://calcofi.github.io/calcofi4r/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
