---
title: "Matching biology to environment: query 75 years of ocean data, anywhere"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.6
vignette: >
  %\VignetteIndexEntry{Matching biology to environment: query 75 years of ocean data, anywhere}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning  = FALSE,
  message  = FALSE,
  comment  = "#>",
  dpi      = 110,
  out.width = "100%")
```

## A new paradigm: 22 GB of CalCOFI, queryable from anywhere

CalCOFI has been sampling the California Current since 1949 — three quarters of
a century of co-located **biological** (net-tow ichthyoplankton + zooplankton
biomass) and **environmental** (CTD-bottle) observations. Until recently, using
that record for cross-domain analysis meant *picking your friction*: a Postgres
server with credentials, an API account, or downloading 22 GB of CSV and
joining it yourself.

It doesn't anymore. Every CalCOFI release is now published as **versioned,
public Apache Parquet** on Google Cloud Storage — and DuckDB can query it
directly over HTTPS. **No credentials. No server. No full download.** DuckDB
reads only the columns and row groups your query actually touches, and it runs
the *same engine* on:

| Where | How |
|---|---|
| **R**, on your laptop | this vignette |
| **Python**, on a notebook server | `import duckdb; con.sql(open('q.sql').read()).df()` |
| **shell**, on the command line | `duckdb < q.sql` |
| **a web browser** | [shell.duckdb.org](https://shell.duckdb.org) — DuckDB-WASM, paste the same SQL, no install |

Below we ask one cross-domain question — *how did Pacific sardine larvae
experience the 2014–2016 marine heatwave?* — answer it with a single
`calcofi4r` call, then unpack what makes that call **reproducible by anyone in
any of those environments**.

## One question, one call, 310 matched observations

`cc_match_ichthyo_by_name()` joins ichthyoplankton tows to CTD-bottle
measurements on the fly: a temporal interval join (within `max_time_hr`) plus a
spatial `ST_Distance_Sphere` filter (within `max_dist_km`), one row per
biological observation with the matched environmental value averaged over the
window. *Relaxed* matching widens the windows to 5 km / 72 hr.

```{r match, cache = TRUE}
library(calcofi4r)
library(dplyr)

d <- cc_match_ichthyo_by_name(
  scientific_name = "Sardinops sagax",
  env_var         = "temperature",
  date_min        = "2014-01-01",
  date_max        = "2019-12-31",
  relax_matching  = TRUE,
  version         = "v2026.05.14")   # pin for archival reproducibility

dim(d)
```

```{r head}
d |>
  select(scientific_name, life_stage, bio_datetime,
         bio_lon, bio_lat, bio_value,
         env_value, env_depth_m, n_env,
         dist_km, time_diff_hr) |>
  head(6)
```

Each row is one net-haul observation of a sardine egg or larva,
joined to the nearest co-located CTD cast that measured temperature. The
`bio_*` columns describe the biology (`bio_value` is the standardized larval
tally, `bio_datetime` / `bio_lon` / `bio_lat` is where + when), the `env_*`
columns describe the matched temperature (`env_value` is the mean of `n_env`
bottle measurements within the match window, `env_depth_m` is the mean
sampling depth), and `dist_km` / `time_diff_hr` are the realized offsets — all
well inside the 5 km / 72 hr relaxed window.

## Where, when, how warm

```{r map, fig.width = 7.2, fig.height = 5.4, fig.cap = "Pacific sardine egg + larva observations across 2014–2019 (Q1–Q4), one point per net haul, coloured by the matched CTD-bottle temperature. 2014–2016 was the Northeast Pacific marine heatwave (the *Blob*) and a strong El Niño — the warmer reds dominate those panels."}
library(ggplot2)
library(sf)

# California Current coastline from Natural Earth (already a Suggest for the pkg)
coast <- rnaturalearth::ne_countries(
  scale = "medium", returnclass = "sf",
  country = c("United States of America", "Mexico"))

d_yr <- d |>
  mutate(year = lubridate::year(bio_datetime)) |>
  sf::st_as_sf(coords = c("bio_lon", "bio_lat"), crs = 4326, remove = FALSE)

ggplot() +
  geom_sf(data = coast, fill = "gray92", colour = "gray70", linewidth = 0.2) +
  geom_sf(data = d_yr,
          aes(colour = env_value, shape = life_stage),
          size = 1.6, alpha = 0.85) +
  scale_colour_viridis_c(
    option = "inferno", direction = 1, name = "Temp\n(°C)") +
  scale_shape_manual(values = c(egg = 16, larva = 17), name = NULL) +
  scale_x_continuous(breaks = c(-124, -120)) +
  coord_sf(xlim = c(-126, -117), ylim = c(30, 38.5), expand = FALSE) +
  facet_wrap(~ year, ncol = 3) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        legend.position  = "right",
        strip.text       = element_text(face = "bold"))
```

## Larvae vs. temperature

```{r scatter, fig.width = 7.2, fig.height = 4.2, fig.cap = "Standardized larval tally (log-scaled) versus matched bottle temperature. A LOESS smoother by life stage suggests the heaviest hauls fall in the cooler 11–14 °C band — typical sardine spawning conditions — with relatively few observations in the warmest waters of the heatwave."}
ggplot(d, aes(env_value, bio_value + 1, colour = life_stage)) +
  geom_jitter(alpha = 0.5, width = 0.0, height = 0.05, size = 1.6) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.9) +
  scale_y_log10() +
  scale_colour_manual(values = c(egg = "#1f78b4", larva = "#e31a1c"),
                      name = NULL) +
  labs(x = "Matched bottle temperature (°C)",
       y = expression(paste("Standardized larval tally + 1  ",
                            "(", log[10], ")"))) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        legend.position  = "top")
```

## The signal of the heatwave, year by year

```{r yearly, results = "asis"}
yearly <- d |>
  mutate(year = lubridate::year(bio_datetime)) |>
  group_by(year) |>
  summarise(
    n_obs        = n(),
    n_egg        = sum(life_stage == "egg"),
    n_larva      = sum(life_stage == "larva"),
    mean_temp_C  = round(mean(env_value), 2),
    range_temp_C = sprintf("%.1f – %.1f",
                           min(env_value), max(env_value)),
    median_tally = round(median(bio_value), 1),
    .groups      = "drop")

knitr::kable(yearly,
  caption = "Annual summary of matched sardine ichthyoplankton + temperature.")
```

```{r yearly-plot, fig.width = 7.2, fig.height = 3.4, fig.cap = "Annual mean matched temperature (top) and number of matched hauls (bottom). 2014–2016 sits above the multi-year mean — the marine heatwave + 2015–16 El Niño — and sardine hauls drop sharply through and after that period."}
library(patchwork)

p1 <- ggplot(yearly, aes(year, mean_temp_C)) +
  geom_col(fill = "#cb181d", width = 0.65) +
  geom_hline(yintercept = mean(yearly$mean_temp_C),
             linetype = "dashed", colour = "gray40") +
  labs(x = NULL, y = "Mean matched\ntemp (°C)") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

p2 <- yearly |>
  tidyr::pivot_longer(c(n_egg, n_larva),
                      names_to = "life_stage", values_to = "n") |>
  mutate(life_stage = sub("n_", "", life_stage)) |>
  ggplot(aes(year, n, fill = life_stage)) +
  geom_col(position = "stack", width = 0.65) +
  scale_fill_manual(values = c(egg = "#1f78b4", larva = "#e31a1c"),
                    name = NULL) +
  labs(x = "Year", y = "Matched hauls") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        legend.position  = "top")

p1 / p2
```

## How `cc_match_*` keeps it portable

Every helper attaches the **exact, fully-interpolated SQL** that produced the
result, plus query metadata, as attributes on the returned data. That is the
package's reproducibility contract:

```{r sql-attrs}
meta <- attr(d, "query_meta")
str(meta)
```

`meta$release_version` pins which release the result came from
(`r meta$release_version` here). `meta$source_urls` is the full list of public
GCS parquet files the query reads. And `attr(d, "sql")` is the literal query —
no `dplyr` translation, no hidden state, ~90 lines of plain DuckDB SQL:

```{r sql-len}
sql <- attr(d, "sql")
length(strsplit(sql, "\n")[[1]])
```

```{r sql-head}
cat(substr(sql, 1, 600), "\n…\n")
```

To get the SQL *without* running it — e.g. to share, save into version
control, or hand to a colleague who works in Python — pass `return_sql = TRUE`:

```r
sql <- cc_match_ichthyo_by_name("Sardinops sagax", return_sql = TRUE)
writeLines(sql, "sardine_match.sql")
```

This is the same artifact the [CalCOFI Integrated
App](https://app.calcofi.io/int) ships in its data-download bundle's `query/`
folder — re-run it anywhere, get identical rows.

## Run it from anywhere

Because the query is plain SQL against public Parquet, the **same string** runs
in any DuckDB client:

**Python**

```python
import duckdb
con = duckdb.connect()
con.sql("INSTALL httpfs; LOAD httpfs; INSTALL spatial; LOAD spatial;")
df = con.sql(open("sardine_match.sql").read()).df()
```

**DuckDB CLI**

```sh
duckdb -c "INSTALL httpfs; LOAD httpfs; INSTALL spatial; LOAD spatial;
$(cat sardine_match.sql)"
```

**A web browser, no install** — paste the same SQL into
[shell.duckdb.org](https://shell.duckdb.org). It runs in DuckDB-WASM, in
JavaScript, against the same public GCS bucket. Try it on a phone if you like.

## Version pinning for archival reproducibility

If your analysis is going into a paper, pin the release:

```r
d <- cc_match_ichthyo_by_name(
  "Sardinops sagax", env_var = "temperature",
  date_min = "2014-01-01", date_max = "2019-12-31",
  relax_matching = TRUE,
  version = "v2026.05.14")
```

Every URL inside `attr(d, "sql")` then carries `v2026.05.14`, so the query
references **immutable Parquet** — a colleague re-running it in 10 years gets
the same rows, even after dozens of new releases have shipped. The release
version, the GCS URLs, the windows, and the join method are all in
`attr(d, "query_meta")`; ship that beside your figures and your analysis is
self-describing.

## Beyond `cc_match_ichthyo_by_name()`

The wrappers cover the most common cases:

- **`cc_match_ichthyo_by_name()`** — filter biology by scientific name (this
  vignette).
- **`cc_match_ichthyo_by_taxon()`** — filter by a WoRMS `worms_id` *and every
  descendant*, resolved by a recursive walk of the taxon tree on GCS. Replaces
  the old API's ITIS-`path`-regex pattern.
- **`cc_match_zooplankton_biomass()`** — match net-tow displacement-volume
  biomass (`net.totalplankton` / `net.smallplankton`) to bottle measurements.

When you need a biological or environmental side the wrappers don't cover (a
custom filter, an extra join, a different grouping), call the core engine
**`cc_match_bio_env(bio, env, ...)`** directly with your own `SELECT`
subqueries — see `?cc_match_bio_env` for the column contract.

For the broader picture — what tables ship in a release, how to query the
hive-partitioned CTD profiles, and the relationship to the Integrated App
download bundle — see the
[Data Access](https://calcofi.io/docs/data-access.html) and
[Matching Helpers](https://calcofi.io/docs/helpers.html) chapters of the
CalCOFI documentation book.
