---
title: "calcofi4r"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{calcofi4r}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning  = FALSE,
  message  = FALSE,
  comment  = "#>")
```

## Introduction

The `calcofi4r` package provides access to the CalCOFI (California Cooperative Oceanic Fisheries Investigations) database, which contains over 75 years of oceanographic and biological data from the California Current ecosystem.

The database is organized into a tidy relational structure stored as Parquet files on Google Cloud Storage. This allows fast queries without downloading the entire database.

## Connect to the Database

```{r connect}
library(calcofi4r)
library(dplyr)
library(DBI)
library(sf)
library(mapview)

# connect to the latest CalCOFI database release
con <- cc_get_db()
q <- dbExecute(con, "INSTALL spatial; LOAD spatial;")

# list available tables
dbListTables(con)
```

## Convenience Functions

The package provides convenience functions for common operations:

```{r convenience}
# list available versions
cc_list_versions()

# list tables
cc_list_tables()

# describe a table
cc_describe_table("obs")

# list measurement types
cc_list_measurement_types() |> head(10)
```

### Read Data Directly

Convenience functions return tibbles with optional filtering:

```{r read-convenience}
# taxonomy: one row per taxon, keyed worms:<id> or itis:<id>
taxa <- cc_read_taxon()
head(taxa)

# ichthyoplankton occurrences
ichthyo_sample <- cc_read_ichthyo() |> head(100)
head(ichthyo_sample)
```

## Database Schema

The database used to expose one table triple per dataset — `ichthyo`, `bottle`,
`bottle_measurement`, `casts`, `species` and roughly thirty more. Those are gone.
Every dataset now projects into one small **core** family, so a query written
against ichthyoplankton works unchanged against CTD, zooplankton or seabirds:

- **`obs`** — the occurrence table, one scalar per row. `realm` is `bio` or
  `env`; biology carries `taxon_key`, environment carries a measurement type.
- **`sample`** — one row per physical sampling event (cast, tow, net, bottle,
  transect), with an adjacency list via `parent_sample_key` / `root_sample_key`.
- **`obs_attribute`** — sub-occurrence detail: length and stage frequency, plus
  categorical behaviour.
- **`sample_measurement`** — event-level effort (volume sampled, haul factor,
  cast conditions).

Shared references: **`taxon`** (one row per taxon, keyed `worms:<id>` or
`itis:<id>`), **`dataset_taxon`** (each dataset's vocabulary crosswalked to it),
**`taxon_group`**, **`cruise`**, **`ship`**, **`grid`**, **`measurement_type`**,
**`spatial`** / **`spatial_attribute`**, **`region`**, **`dataset`**, **`lookup`**.

Full column-level reference: [calcofi.io/schema](https://calcofi.io/schema/).

```{r schema}
# show row counts for each table
tables <- dbListTables(con)
tibble(
  table = tables,
  rows  = sapply(tables, function(t) {
    dbGetQuery(con, sprintf("SELECT COUNT(*) as n FROM %s", t))$n
  })) |>
  arrange(desc(rows))
```

## Query Environmental Data

Environmental measurements live in `obs` alongside biology, distinguished by
`realm`, in a long format: one row per (sample, depth, measurement type).
Position and time come from `sample`.

```{r bottle-temp}
# surface temperature with location and depth
d_temp <- dbGetQuery(con, "
  SELECT
    s.longitude AS lon,
    s.latitude  AS lat,
    s.datetime,
    o.depth_min_m AS depth_m,
    o.measurement_value AS temperature
  FROM obs o
  JOIN sample s USING (sample_key)
  WHERE o.measurement_type = 'temperature'
    AND o.measurement_value IS NOT NULL
    AND o.depth_min_m <= 10
  LIMIT 100000")

head(d_temp)
nrow(d_temp)
```

### Summarize by Location

```{r temp-summary}
# summarize surface temperature by location
d_t <- d_temp |>
  group_by(lon, lat) |>
  summarize(
    n     = n(),
    t_avg = mean(temperature, na.rm = TRUE),
    .groups = "drop") |>
  filter(!is.na(lon), !is.na(lat)) |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE)

head(d_t)
```

## CalCOFI Grid

The CalCOFI sampling grid defines standard station positions. The package includes pre-loaded grid data:

- `cc_grid` - station polygons
- `cc_grid_ctrs` - station centroids
- `cc_grid_zones` - aggregated zones by station pattern

```{r grid}
# show the CalCOFI grid colored by zone
mapview(cc_grid, zcol = "zone_key", layer.name = "Zone") +
  mapview(cc_grid_ctrs, cex = 1, col.regions = "black", legend = FALSE)
```

### Grid from Database

The grid is also available in the database with additional attributes:

```{r grid-db}
# query grid from database (includes geometry)
grid_db <- dbGetQuery(con, "SELECT * EXCLUDE(geom, geom_ctr) FROM grid")
head(grid_db)
```

## Show Effort by Grid Cell

Join temperature observations to the CalCOFI grid to show sampling effort:

```{r effort-grid}
# count observations per grid cell
n_grid <- cc_grid |>
  st_join(d_t) |>
  group_by(sta_key) |>
  summarize(n = sum(n, na.rm = TRUE))

mapview(n_grid, zcol = "n", layer.name = "Observations")
```

### Show Effort by Station Point

```{r effort-points}
# join counts to centroids
n_pts <- cc_grid_ctrs |>
  left_join(
    n_grid |> st_drop_geometry() |> select(sta_key, n),
    by = "sta_key")

mapview(n_pts, cex = "n", layer.name = "Observations")
```

## Map Contours

Interpolate temperature data to create contour maps using Inverse Distance Weighting (IDW).

### All Zones

```{r contour-all}
# interpolate points to raster using IDW
r_all <- pts_to_rast_idw(d_t, "t_avg", cc_grid_zones)

# generate contour polygons
p_all <- rast_to_contours(r_all, cc_grid_zones)
mapview(p_all, zcol = "z_avg", layer.name = "Temp (C)")
```

### Standard and Extended Pattern

```{r contour-ext}
# filter to standard + extended zones
aoi_ext <- cc_grid_zones |>
  filter(sta_pattern %in% c("standard", "extended"))

# interpolate and contour
r_ext <- pts_to_rast_idw(d_t, "t_avg", aoi_ext)
p_ext <- rast_to_contours(r_ext, aoi_ext)
mapview(p_ext, zcol = "z_avg", layer.name = "Temp (C)")
```

## Query Ichthyoplankton Data

The ichthyoplankton survey counts fish larvae by species across sampling sites.

```{r ichthyo}
# top 10 taxa by total count. `taxon_key` is global, so the same join works for
# any dataset — swap the dataset_key and this counts euphausiids or seabirds.
top_species <- dbGetQuery(con, "
  SELECT
    t.scientific_name,
    t.common_name,
    SUM(o.measurement_value)      AS total_count,
    COUNT(DISTINCT o.sample_key)  AS n_samples
  FROM obs o
  JOIN taxon t USING (taxon_key)
  WHERE o.realm = 'bio'
    AND o.dataset_key = 'swfsc_ichthyo'
  GROUP BY 1, 2
  ORDER BY total_count DESC
  LIMIT 10")

top_species
```

### Species Distribution

Map the distribution of a common species:

```{r species-map}
# get Northern Anchovy observations with locations
# `sample` replaces the old site -> tow -> net chain: obs joins it directly.
anchovy <- dbGetQuery(con, "
  SELECT
    s.latitude  AS lat,
    s.longitude AS lon,
    SUM(o.measurement_value) AS count
  FROM obs o
  JOIN taxon  t USING (taxon_key)
  JOIN sample s USING (sample_key)
  WHERE t.scientific_name = 'Engraulis mordax'
    AND o.realm = 'bio'
  GROUP BY 1, 2") |>
  filter(!is.na(lon), !is.na(lat)) |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

mapview(anchovy, cex = "count", layer.name = "Anchovy count")
```

## Cruise Timeline

View the temporal coverage of CalCOFI cruises:

```{r cruises}
# get cruise timeline (date_ym is YYYYMM format)
cruises <- dbGetQuery(con, "
  SELECT
    cruise_key,
    CAST(SUBSTRING(CAST(date_ym AS VARCHAR), 1, 4) AS INTEGER) as year,
    CAST(SUBSTRING(CAST(date_ym AS VARCHAR), 5, 2) AS INTEGER) as month,
    ship_key
  FROM cruise
  WHERE date_ym IS NOT NULL
  ORDER BY date_ym")

# count cruises by year
cruises |>
  count(year) |>
  filter(!is.na(year)) |>
  ggplot2::ggplot(ggplot2::aes(year, n)) +
  ggplot2::geom_col(fill = "steelblue") +
  ggplot2::labs(
    title = "CalCOFI Cruises by Year",
    x = "Year", y = "Number of Cruises") +
  ggplot2::theme_minimal()
```

## Available Measurement Types

`measurement_type` is the shared vocabulary across every dataset, with units and
the physical bounds each value is validated against:

```{r measurements}
# list all measurement types
dbGetQuery(con, "
  SELECT measurement_type, description, units
  FROM measurement_type
  ORDER BY measurement_type") |>
  head(20)
```

## Disconnect

Always close the database connection when finished:

```{r disconnect}
dbDisconnect(con)
```

## Package Data Objects

The package also includes pre-loaded spatial data objects for convenience:

- `cc_grid` - CalCOFI station grid polygons (sf)
- `cc_grid_ctrs` - Station centroids (sf)
- `cc_grid_zones` - Aggregated zone polygons (sf)
- `cc_bottle` - Sample bottle data for examples (tibble)

See also `vignette("bio-env-matching")` for joining biology to environment, and
`vignette("ctd-temperature-anomalies")` for transects, climatologies and
anomalies.

These can be used without connecting to the database for quick spatial operations.
