---
title: "calcofi4r"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{calcofi4r}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = F,
  message = F,
  comment = "#>"
)
# devtools::install_local(here::here(), force=T)
# devtools::load_all()
```

## Show Effort

The CalCOFI grid (`cc_grid`) and centroids (`cc_grid_ctrs`) show the  [CalCOFI Station Positions – CalCOFI](https://calcofi.org/sampling-info/station-positions/) at varying separation distances of station positions (`sta_dpos`) in the CalCOFI coordinate system from nearshore (`5`), to offshore (`10`) to outside the 113 extended station pattern to historically visited areas (`20`).

This allows summarization of effort within each station and defining a study area according to these criteria (e.g., `sta_dpos`).

### Grid by Zone

```{r}
library(calcofi4r)
library(dplyr)
library(mapview)
library(sf)
library(stars)

mapview(cc_grid, zcol="zone_key") +
  mapview(cc_grid_ctrs, cex = 1)
```

### Summarize Temperature from Bottle Data

```{r}
# show first 6 rows of example data frame with temperature in space as variable
head(cc_bottle)

# summarize temperature by location and convert to points
d_t <- cc_bottle %>% 
  filter(
    !is.na(t_degc)) %>% 
  group_by(lon, lat) %>% 
  summarize(
    n = n(),
    v = mean(t_degc),
    .groups = "drop") %>% 
  st_as_sf(
    coords = c("lon", "lat"), remove = F,
    crs = 4326)
```


### Show Effort by Grid Cell

```{r}
n_ply <- cc_grid %>% 
  st_join(d_t) %>% 
  group_by(sta_key) %>% 
  summarize(
    n = sum(n))
mapview(n_ply, zcol="n", layer.name = "n")
```

### Show Effort by Grid Point

```{r}
n_pts <- cc_grid_ctrs %>% 
  left_join(
    n_ply %>% 
      st_drop_geometry() %>% 
      select(sta_key, n),
    by = "sta_key")
mapview(n_pts, cex="n", layer.name = "n")
```

## Map contours 

You can map the contours of input data using the faster, rougher Inverse-Distance Weighting (IDW; `pts_to_rast_idw()`, `rast_to_contours()`) technique or the slower, smoother Generalized Additive Model (GAM; `pts_to_contours_gam()`).

### All Zones

```{r pts_to_rast_idw}
# interpolate points to raster using IDW
r_k_all <- pts_to_rast_idw(d_t, "v", cc_grid_zones)
mapView(st_as_stars(r_k_all))
```

```{r}
# generate contour polygons from raster
p_k_all <- rast_to_contours(r_k_all, cc_grid_zones)
mapView(p_k_all, zcol = "z_avg")
```

### Standard and Extended Station Pattern Zones

```{r}
aoi_ext <- cc_grid_zones |> 
  filter(
    sta_pattern %in% c("standard", "extended"))

# interpolate points to raster using IDW
r_k_ext <- pts_to_rast_idw(d_t, "v", aoi_ext)
mapView(st_as_stars(r_k_ext), layer.name="Temp.(C)")
```

```{r}
# generate contour polygons from raster
p_k_ext <- rast_to_contours(r_k_ext, aoi_ext)
mapView(p_k_ext, zcol = "z_avg", layer.name="Temp.(C)")
```

### Standard Nearshore Zone

```{r}
aoi_stdnr <- cc_grid_zones %>% 
  filter(
    sta_pattern  == "standard",
    sta_shore    == "nearshore")

# interpolate points to raster using IDW
r_k_stdnr <- pts_to_rast_idw(d_t, "v", aoi_stdnr)
mapView(st_as_stars(r_k_stdnr), layer.name="Temp.(C)")
```

```{r}
# generate contour polygons from raster
p_k_stdnr <- rast_to_contours(r_k_stdnr, aoi_stdnr)
mapView(p_k_stdnr, zcol = "z_avg", layer.name="Temp.(C)")
```

<!--## Map contours using a Generalized Additive Model (GAM)

Map contours of an oceanographic variable from CTD cast bottle data. The `pts_to_contours_gam()` method uses a bivariate smoothing function on longitude and latitude with a generalized additive model (GAM).

### All Zones

```{r pts_to_contours_gam, eval=F, echo=F}
# calculate map contours
p_g_all <- pts_to_contours_gam(d_t, cc_grid_zones)

# show map contours
mapView(p_g_all, zcol="v", layer.name="Temp.(C)")
```

### Standard and Extended Station Pattern Zones

```{r, eval=F, echo=F}
# calculate map contours
p_g_ext <- pts_to_contours_gam(d_t, aoi_ext)

# display map contours
mapView(p_g_ext, zcol="v", layer.name="Temp.(C)")
```

### Standard Nearshore Zone

```{r, eval=F, echo=F}
# calculate map contours
p_g_stdnr <- pts_to_contours_gam(d_t, aoi_stdnr)

# display map contours
mapView(p_g_stdnr, zcol="v", layer.name="Temp.(C)")
```
-->

## Plot Time Series

Plot time series of an oceanographic variable from CTD cast bottle data.

```{r plot_timeseries}
# get variables
(v <- get_variables())

# fetch time series data for the first variable from CalCOFI API
(d <- get_timeseries("ctd_bottles.t_degc"))

# plot time series with the first variable
with(v[1,],
  plot_timeseries(
    # data and columns (from d)
    d, year, val_avg, val_sd,
    # plot attributes (from v)
    plot_title, plot_label, plot_color))
```

## Plot Depth

Plot depth of an oceanographic variable from CTD cast bottle data.

```{r plot_depth}
head(bottle_temp_depth)

# plot depth with the station data
plot_depth(df = bottle_temp_depth, variable = "Temperature")
```

<!--
## Map Raster

Map raster of interpolated oceanographic variable for a cruise.
-->
```{r map_raster, eval=F, echo=F}
# get cruises
(z <- get_cruises())

# get path of temporary file to store raster
(r_tif <- tempfile(fileext=".tif"))

# use second variable from previously fetched v
c(v$table_field[2], v$plot_label[2])

# fetch interpolated raster from CalCOFI API
get_raster(
  variable = v$table_field[2],
  cruise_id = "2020-01-05-C-33RL",
  depth_m_min = 0, depth_m_max = 200,
  out_tif = r_tif)

# cleanup by deleting temporary file
unlink(r_tif)

# read raster
r <- raster::raster(r_tif)

# plot raster
map_raster(r, v$plot_label[2])
```
