---
title: "Citing CalCOFI data"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.6
vignette: >
  %\VignetteIndexEntry{Citing CalCOFI data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning  = FALSE,
  message  = FALSE,
  comment  = "#>")
```

## The release cites itself; every dataset in it cites its own source

CalCOFI's integrated database is not one thing to cite — it is a **release** (a
specific, versioned snapshot) built from **datasets**, each contributed by its
own program under its own license, and often its own DOI. Citing "CalCOFI" and
citing the CTD-bottle time series and citing the Farallon seabird census are
three different citations, and a paper that used all three owes all three.

Every field involved — `citation_main`, `license`, `doi`,
`acknowledgement` — is checked at release time
(`calcofi4db::check_dataset_citation()`, the attribution contract): a citation
with a year and a locator, a license registered in `metadata/license.csv`, a
DOI that actually resolves. `cc_cite()` is the one place that reads those
fields back out and formats them, so you never have to build a citation string
by hand or guess whether a `license_url` is required.

```{r cite-all}
library(calcofi4r)

REL <- cc_latest_version()
cc_cite(version = REL)
```

The **first** entry is always the release itself — *CalCOFI Integrated
Database, release `r REL`* — cited under the concept DOI once Zenodo has
minted one for the tag, or a stable `db-schema` URL until then. Every entry
after it is one dataset, in alphabetical `dataset_key` order by default.

## Citing just what you used

Pass the `dataset_key`(s) you actually queried — never everything, and never
whichever ones you remember typing.

```{r cite-some}
cc_cite(c("calcofi_bottle", "calcofi_ctd-cast"), version = REL)
```

Better still, hand `cc_cite()` the **query result itself**. Any tibble
carrying a `dataset_key` column (`cc_read_obs()`, `cc_read_sample()`, a
`cc_match_*()` join, your own SQL) works directly — the distinct keys the
query actually touched are what gets cited, in the order they first appear:

```{r cite-query}
library(dplyr)

d <- cc_read_obs(
  datasets = c("calcofi_dic", "calcofi_bottle"),
  measurement_types = "temperature",
  version = REL) |>
  head(500)

cc_cite(d, version = REL)
```

An unrecognized `dataset_key` is an error naming it, rather than a citation
silently missing a dataset:

```{r cite-unknown, error = TRUE}
cc_cite("not_a_real_dataset", version = REL)
```

## Three formats

`format = "text"` (the default, above) is meant for a methods section or an
email. `format = "bibtex"` builds a `.bib`-ready entry per citation, and
`format = "csl"` returns [CSL-JSON](https://citeproc-js.readthedocs.io/) items
— what Zotero, Pandoc citations and most reference managers import directly.

```{r cite-bibtex, results = "asis"}
cat(cc_cite("calcofi_dic", version = REL, format = "bibtex"))
```

```{r cite-csl}
str(cc_cite("calcofi_dic", version = REL, format = "csl")[[1]])
```

`format = "bibtex"` builds every entry **offline**, from the fields already on
the release's `dataset` table and in its `catalog.json` — nothing here calls
the network by default. `resolve = TRUE` is the only network path: for any
entry with a DOI, it fetches the DOI's own BibTeX from `doi.org` instead
(falling back to the offline entry if that fetch fails), which some reference
managers format slightly differently.

```{r cite-resolve, eval = FALSE}
cc_cite("calcofi_dic", version = REL, format = "bibtex", resolve = TRUE)
```

## Citing an older release

A release frozen before the attribution contract landed (2026-09-03) has no
`citation` in its `catalog.json`. `cc_cite()` computes the same wording
`calcofi4db::release_citation()` would have written rather than erroring, and
marks the result so you can tell which happened:

```{r cite-source}
attr(cc_cite(character(0), version = REL), "source")
```

`"release"` means the catalog carried its own citation; `"computed"` means
`cc_cite()` derived it.

## Citing the software, separately

`cc_cite()` is for the **data**. To cite the R package itself — appropriate
alongside the data citation when `calcofi4r` did real analytical work, not
just I/O — use base R's own mechanism, which reads `DESCRIPTION`:

```{r cite-software}
citation("calcofi4r")
```

The Python sibling, [`calcofi4py`](https://calcofi.io/calcofi4py/), mirrors
every part of `cc_cite()` above byte-for-byte under the same name
(`calcofi4py.cc_cite()`) — the same release and dataset citations, read the
same way, for anyone working in a notebook instead of R.
