---
title: "Summer CTD temperature anomalies against a 1993-2013 baseline"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.6
vignette: >
  %\VignetteIndexEntry{Summer CTD temperature anomalies against a 1993-2013 baseline}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  warning   = FALSE,
  message   = FALSE,
  comment   = "#>",
  dpi       = 110,
  out.width = "100%")
```

## What this shows

A complete anomaly analysis using five `calcofi4r` functions —
`cc_transect_stations()`, `cc_transect_section()`, `cc_climatology()`,
`cc_anomaly()` and `cc_transect_matrix()` — against the public CalCOFI release,
with no credentials and no SQL.

The worked question is a real one, from Rasmus Swalethorp (SIO) for a CCIEA
workgroup: *extract the summer cruise temperature data and generate anomaly
plots using 1993–2013 as the baseline.*

**1993–2013 is newly possible.** Until the August 2026 release the 1 m-binned
CTD record jumped straight from **1998 to 2003** — calcofi.org publishes *Final*
1 m-binned products for 1998 and 2003 onward, and only raw cast files before
that. A backfill from Jim Wilkinson's archive closed the gap, so that baseline
now means what it says instead of quietly meaning "1998, plus 2003–2013". The
last section quantifies what the difference is worth.

```{r libs}
library(calcofi4r)
library(dplyr)
library(ggplot2)
library(tidyr)

con <- cc_get_db()
```

### Parameters, stated once

The choice of baseline is the single assumption a reader of an anomaly plot most
needs to see, so it lives at the top rather than inside a call.

```{r knobs}
BASELINE  <- c(1993, 2013)   # inclusive
SUMMER    <- 7:8             # CalCOFI's summer occupation runs Jul-Aug
LINE      <- 93.3            # the line off San Diego; every core line works
VAR       <- "temperature_ave"
DEPTH_MAX <- 500
MIN_N     <- 3               # baseline cells thinner than this are dropped
```

## One transect

A transect here is **one CalCOFI line, ordered by station number ascending** —
nearshore to offshore. There are no endpoints to pick, which is what makes any
line × cruise drawable without user input. It is deliberately *not* the order
stations were occupied: that is the ship's track, so its direction is whichever
way the ship steamed, and it is missing on roughly half the release's cast rows.

```{r one-section}
cruises <- tbl(con, "sample") |>
  filter(dataset_key == "calcofi_ctd-cast", sample_type == "cast") |>
  mutate(yr = as.integer(strftime(datetime, "%Y")),
         mo = as.integer(strftime(datetime, "%m"))) |>
  count(cruise_key, yr, mo, data_stage, name = "n_casts") |>
  collect()

summer <- filter(cruises, mo %in% SUMMER)
cruise_recent <- summer |> filter(yr == max(yr)) |> slice(1) |> pull(cruise_key)

sta <- cc_transect_stations(con, LINE, cruise_recent)
sec <- cc_transect_section(
  con, LINE, cruise_recent, variables = VAR, depth_max = DEPTH_MAX)

head(sec)
```

`cc_transect_stations()` also decides the horizontal ruler. `x = "occupied"`
(the default) measures between the stations this cruise actually occupied, so
the section fills the plot; `x = "line"` measures along the full line, so cruises
covering different extents are comparable in width.

That is not cosmetic. Line 93.3 has not been sampled past station 90 since
January 2025, though 113 of the 130 cruises before that reached station 120:

```{r rulers}
occ <- cc_transect_stations(con, LINE, cruise_recent, x = "occupied")
lin <- cc_transect_stations(con, LINE, cruise_recent, x = "line")

c(occupied_km = max(occ$dist_km), along_line_km = max(lin$dist_km))
```

## The baseline, and a filter before it

```{r clim}
clim <- cc_climatology(
  con, variables = VAR, years = BASELINE, depth_max = DEPTH_MAX, min_n = MIN_N)

attr(clim, "baseline")
nrow(clim)
```

`cc_climatology()` is a plain mean per (station, 5 m depth bin, calendar month) —
the finest grouping CalCOFI's design supports, since quarterly-ish cruises over
decades give many *years* per calendar month at a station but only a handful of
days. It returns `clim_n` so a thin cell is visible rather than silently trusted.

### Screen the source for physically impossible values

**Do this before computing a baseline, not after looking at the result.** The
released CTD carries a small number of surface readings that are not ocean
temperatures — the classic soak artifact, where the sensor is still warm from
sitting on deck and has not equilibrated in the first metre or two of the
downcast.

There are only 18 such rows in 13.5 million, and the release's declared bounds
(−2 to 40 °C, deliberately set to "impossible" rather than "unusual") let them
through. But they land in the sparsely-sampled northern lines, where an April
baseline cell may hold only 2–7 observations in the entire 1993–2013 window — so
one 38 °C reading moves that cell's mean by more than 10 °C, and every other
April cruise at that station then reads as a spurious ~11 °C **cold** anomaly.

The data makes the cut obvious: there is nothing at all between 26 and 35 °C.

```{r outlier-gap}
temp_all <- tbl(con, "obs") |>
  filter(dataset_key == "calcofi_ctd-cast", measurement_type == VAR,
         !is.na(measurement_value)) |>
  select(measurement_value, depth_min_m) |>
  collect()

temp_all |>
  mutate(band = cut(measurement_value, c(-Inf, 24, 26, 30, 35, Inf),
                    labels = c("<24", "24-26", "26-30", "30-35", ">=35"))) |>
  count(band)
```

So a regional ceiling well inside that gap removes every artifact and nothing
else. This is a *client-side* screen — the release keeps the rows, and the
values are reported to the data provider as
[Q22](https://calcofi.io/workflows/ingest_calcofi_ctd-cast.html#questions-for-data-providers)
so they can be fixed at source rather than papered over by every downstream user
in their own way.

```{r screen}
# Two-sided, on purpose. An earlier draft screened only the upper bound, because
# the artifact I had in front of me was a hot one — which would have sailed past
# the cold artifact that motivated this whole exercise (a failed sensor averaged
# into TempAve produced values near -47 degC). A screen that can only catch the
# failure you already know about is not a screen.
#
# 0 and 30 degC: roughly 5 degC outside the coldest and warmest genuine CalCOFI
# readings, and far outside anything the region produces. Nothing legitimate is
# near either bound.
TEMP_RANGE_REGIONAL <- c(0, 30)

screen <- function(d, col = "measurement_value")
  filter(d, .data[[col]] >= TEMP_RANGE_REGIONAL[1],
            .data[[col]] <= TEMP_RANGE_REGIONAL[2])

clim_screened <- cc_climatology(
  con, variables = VAR, years = BASELINE,
  depth_max = DEPTH_MAX, min_n = MIN_N) |>
  filter(clim_mean >= TEMP_RANGE_REGIONAL[1],
         clim_mean <= TEMP_RANGE_REGIONAL[2])

nrow(clim) - nrow(clim_screened)   # baseline cells the screen removes
```

## Anomalies for one section

`cc_anomaly()` differences a section against the climatology, matched on station,
calendar month and depth bin. A cell with no baseline comes back `NA`, **never
0** — an unsampled baseline is not a zero anomaly, and collapsing the two is how
a plot ends up colouring "normal" somewhere never measured.

```{r anomaly}
anom <- cc_anomaly(sec, clim_screened, sta)

round(100 * mean(!is.na(anom$anomaly)))   # % of this section with a baseline
range(anom$anomaly, na.rm = TRUE)
```

```{r fig-section, fig.alt = "Cross-shelf temperature anomaly section"}
lim <- max(abs(anom$anomaly), na.rm = TRUE)

ggplot(anom, aes(dist_km, depth_m, fill = anomaly)) +
  geom_raster(interpolate = TRUE) +
  scale_fill_gradient2(
    low = "#0d366b", mid = "#f0efec", high = "#7d1b28",
    midpoint = 0, limits = c(-lim, lim), na.value = "transparent",
    name = "Δ °C") +
  scale_y_reverse(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(x = "Distance offshore (km)", y = "Depth (m)",
       title = paste0("Line ", LINE, " · ", cruise_recent,
                      " · temperature anomaly")) +
  theme_minimal(base_size = 11)
```

For a heatmap that wants a matrix rather than a long table,
`cc_transect_matrix()` pivots to `z[[depth]][[station]]` with `x` the station
distances — the shape a Plotly or ODV-style section takes directly.

```{r matrix}
m <- cc_transect_matrix(anom, value = "anomaly")
str(m, max.level = 1)
```

## All summer cruises: an anomaly time series

The headline figure. Rather than sectioning every cruise, difference **every**
summer observation against the same baseline and average by depth layer.

Two choices worth stating, because both are ways this goes quietly wrong:

- **Average the anomalies, not the values.** Averaging temperatures first would
  let a year that happened to sample more nearshore stations read as cold purely
  from where it sampled. Differencing per station and depth removes each
  station's own climate before any averaging.
- **Only cells with a real baseline count.** A station newly added to the grid
  has no 1993–2013 history; treating that as a zero anomaly drags every mean
  toward zero by an amount nobody can see.

```{r series}
obs_summer <- tbl(con, "obs") |>
  filter(dataset_key == "calcofi_ctd-cast", measurement_type == VAR,
         !is.na(measurement_value), !is.na(depth_min_m), !is.na(datetime),
         depth_min_m <= DEPTH_MAX) |>
  mutate(yr  = as.integer(strftime(datetime, "%Y")),
         mon = as.integer(strftime(datetime, "%m"))) |>
  filter(mon %in% SUMMER) |>
  mutate(depth_m = round(depth_min_m / 5) * 5) |>
  select(grid_key, cruise_key, yr, mon, depth_m,
         variable = measurement_type, value = measurement_value) |>
  collect() |>
  screen("value")            # the same client-side screen, applied to the data

anom_summer <- obs_summer |>
  inner_join(clim_screened,
             by = c("grid_key", "mon" = "month", "depth_m", "variable")) |>
  mutate(anomaly = value - clim_mean)

series <- anom_summer |>
  mutate(layer = cut(depth_m, c(-1, 50, 100, 200, 500),
                     labels = c("0-50 m", "50-100 m", "100-200 m", "200-500 m"))) |>
  filter(!is.na(layer)) |>
  summarize(anomaly = mean(anomaly), n = n(), .by = c(yr, layer))

head(series)
```

A sanity check before reading anything into it: the baseline years must sit near
zero by construction, and the series should reproduce events we already know
happened. It does — 2014-15 is the marine heatwave ("the Blob"), 2016 its El
Niño tail, and 2026 is warm at *every* depth rather than just at the surface.
2018 is absent because there was no summer occupation that year.

```{r fig-series, fig.alt = "Summer temperature anomaly by year and depth layer"}
ggplot(series, aes(yr, anomaly, colour = layer)) +
  annotate("rect", xmin = BASELINE[1] - .5, xmax = BASELINE[2] + .5,
           ymin = -Inf, ymax = Inf, fill = "grey70", alpha = .16) +
  geom_hline(yintercept = 0, linewidth = .4, colour = "grey40") +
  geom_line(linewidth = .7) +
  geom_point(size = 1.6) +
  scale_colour_manual(values = c("#e34948", "#3987e5", "#256abf", "#0d366b"),
                      name = NULL) +
  labs(x = NULL, y = "Temperature anomaly (°C)",
       title = paste0("CalCOFI summer CTD temperature anomaly, ",
                      BASELINE[1], "-", BASELINE[2], " baseline"),
       caption = "Grey band marks the baseline window") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

## What the backfill is worth

The same analysis against a 1998–2013 baseline — what would have been possible
before the 1993–2002 cruises were ingested.

```{r sensitivity}
clim_98 <- cc_climatology(
  con, variables = VAR, years = c(1998, 2013),
  depth_max = DEPTH_MAX, min_n = MIN_N) |>
  filter(clim_mean >= TEMP_RANGE_REGIONAL[1],
         clim_mean <= TEMP_RANGE_REGIONAL[2])

cmp <- obs_summer |>
  inner_join(clim_98, by = c("grid_key", "mon" = "month", "depth_m", "variable")) |>
  mutate(anomaly_98 = value - clim_mean) |>
  filter(depth_m <= 50) |>
  summarize(anomaly_98 = mean(anomaly_98), .by = yr)

comparison <- series |>
  filter(layer == "0-50 m") |>
  select(yr, anomaly_9313 = anomaly) |>
  inner_join(cmp, by = "yr") |>
  mutate(difference = anomaly_9313 - anomaly_98)

summary(abs(comparison$difference))
```

## Reading it honestly

**The baseline is 21 years, not a 30-year WMO normal.** CalCOFI's 1 m-binned CTD
record does not reach back far enough for one without splicing in ship hydrocast
data of a different vintage. 1993–2013 does span both the 1997–98 El Niño and the
1998–99 La Niña, and it ends before the 2014–16 marine heatwave — so the heatwave
and everything since read as departures rather than being folded into the normal.

**Coverage is not constant.** A spatial mean over "the line" covers different
ocean in different years. The per-station differencing above removes each
station's own climate, which is most of the problem — but a year that skipped
the offshore stations still has no offshore signal in it.

**Preliminary cruises are fine for temperature, and not for the corrected
series.** Temperature is not bottle-corrected, so a `preliminary_without_bottle`
cruise carries the temperature it always will. Salinity, oxygen and chlorophyll
on those cruises are a different matter: the corrected forms do not exist until
the bottle merge runs, and the uncorrected sensor series (`salinity_1`,
`oxygen_ml_l_1`) are what is available.

**The two-sensor averages are recomputed during ingest, not taken as shipped.**
The source CSVs carry pre-computed `TempAve`, `SaltAve_Corr`, `OxAve_StaCorr` and
`OxAveuM_StaCorr`, and where one sensor is missing or has failed those sometimes
fold the bad sensor in. The obvious form averages in the `-99` missing marker;
the dangerous form does not — a failed sensor reading −18 °C beside a good one at
20 °C yields an average of 0.77 °C, which is inside any plausible range and worth
an 18 °C false cold anomaly. The ingest treats a sensor outside its declared
bounds as absent and averages over what remains. Reading the source CSVs directly
gives you the unrepaired values; see
[Q21](https://calcofi.io/workflows/ingest_calcofi_ctd-cast.html#questions-for-data-providers).

```{r cleanup, include = FALSE}
DBI::dbDisconnect(con, shutdown = TRUE)
```
